---
title: "Practice 1 — Regularization: Ridge, Lasso vs OLS"
subtitle: "PhD Seminar: Machine Learning in Time Series"
author: "Prof. Marcelo Villena Ch., PhD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    smooth-scroll: true
    code-fold: true
    code-tools: false
    highlight-style: github
    html-math-method: mathjax
execute:
  warning: false
  message: false
  eval: true
---

## Overview

This practice session is divided into two parts. Both use **US CPI inflation data** retrieved directly from FRED and apply an **AR(12)** structure --- that is, 12 monthly lags as predictors --- to compare three forecasting models:

| Model | Penalty                             | Key idea                                           |
|--------------------|-------------------------|----------------------------|
| OLS   | None                                | Standard least squares; can overfit with many lags |
| Ridge | L2 (shrinks all coefficients)       | Reduces variance, retains all predictors           |
| Lasso | L1 (sets some coefficients to zero) | Performs variable selection automatically          |

------------------------------------------------------------------------

## Part A --- Static Train/Test Split

### Objective

Fit OLS, Ridge, and Lasso on a fixed **80/20 temporal split** and compare their out-of-sample RMSE. Ridge and Lasso use a **time-ordered validation** scheme (no random k-fold) to respect the temporal structure of the data.

### Setup

Load the required libraries. If any package is missing, install it first with `install.packages("package_name")`.

```{r, echo=FALSE}
#| label: setup-a
library(quantmod)
library(glmnet)
library(ggplot2)
library(tidyr)
library(dplyr)
library(jtools)
```

### Helper Function: Time-Ordered Tuning

Standard cross-validation randomly shuffles observations, which **leaks future information** into the past. The function below avoids this by holding out the last `val_frac` of the training sample as a temporal validation block.

```{r}
#| label: helper-tune
tune_glmnet_time_split <- function(X, y, alpha, val_frac = 0.20, min_val = 12,
                                   standardize = TRUE) {
  n <- nrow(X)
  v <- max(min_val, floor(val_frac * n))
  if (n <= v + 5) stop("Not enough observations to create a validation split.")

  idx_tr <- 1:(n - v)
  idx_va <- (n - v + 1):n

  # Build a lambda grid from early training only
  fit0 <- glmnet(X[idx_tr, , drop = FALSE], y[idx_tr],
                 alpha = alpha, standardize = standardize)
  lambda_grid <- fit0$lambda

  # Evaluate each lambda on the held-out temporal block
  fit_grid <- glmnet(X[idx_tr, , drop = FALSE], y[idx_tr],
                     alpha = alpha, lambda = lambda_grid,
                     standardize = standardize)

  pred_va  <- predict(fit_grid, s = lambda_grid, newx = X[idx_va, , drop = FALSE])
  y_va_mat <- matrix(y[idx_va], nrow = v, ncol = length(lambda_grid))
  rmse     <- sqrt(colMeans((pred_va - y_va_mat)^2))
  lambda_best <- lambda_grid[which.min(rmse)]

  # Refit on the full training sample using the selected lambda
  fit_full <- glmnet(X, y, alpha = alpha, lambda = lambda_grid,
                     standardize = standardize)

  list(fit = fit_full, lambda = lambda_best, rmse_path = rmse, v = v)
}
```

### Step 1 --- Retrieve Inflation Data

We download the **US Core CPI (CPILFESL)** from FRED and compute **year-over-year inflation** as a percentage change over 12 months.

```{r}
#| label: data-download
cat("Retrieving US CPI data from FRED...\n")
#getSymbols("CPILFESL", src = "FRED")
CPILFESL <- readRDS("../data/CPILFESL.rds")

inflation <- ROC(CPILFESL, n = 12, type = "discrete") * 100
inflation <- na.omit(inflation)
colnames(inflation) <- "Inflation"

plot(inflation, main = "US Year-over-Year Inflation (%)")
```

### Step 2 --- Build the AR(12) Feature Matrix

The `embed()` function stacks lagged values row by row. The result is:

-   **Y**: the inflation value at time $t$\
-   **X**: the 12 preceding monthly values ($t-1, \ldots, t-12$)

```{r}
#| label: feature-matrix
p <- 12
y_vec <- as.numeric(inflation)
N     <- length(y_vec)

lagged_matrix <- embed(y_vec, p + 1)
Y <- lagged_matrix[, 1]
X <- lagged_matrix[, 2:(p + 1)]
colnames(X) <- paste0("Lag_", 1:p)

dates <- index(inflation)[(p + 1):N]
```

### Step 3 --- Temporal Train/Test Split (80/20)

We split the data **chronologically**: the first 80% is used for training, and the last 20% is reserved for out-of-sample evaluation. No observation from the future is used to fit any model.

```{r}
#| label: split
split_index <- floor(nrow(X) * 0.8)

x_train <- X[1:split_index, , drop = FALSE]
y_train <- Y[1:split_index]

x_test     <- X[(split_index + 1):nrow(X), , drop = FALSE]
y_test     <- Y[(split_index + 1):length(Y)]
test_dates <- dates[(split_index + 1):length(dates)]
```

### Step 4 --- Fit OLS (Standard AR(12))

OLS estimates all 12 lag coefficients without any penalty. With many predictors, this can lead to **overfitting**, especially in small samples.

```{r}
#| label: ols
train_df  <- data.frame(y = y_train, x_train)
test_df   <- data.frame(x_test)

ols_model <- lm(y ~ ., data = train_df)
ols_pred  <- predict(ols_model, newdata = test_df)

summ(ols_model)
```

### Step 5 --- Fit Ridge (L2 Penalty)

Ridge shrinks all coefficients toward zero proportionally. The regularization strength $\lambda$ is selected via time-ordered validation (`alpha = 0`).

```{r}
#| label: ridge
ridge_tuned <- tune_glmnet_time_split(x_train, y_train, alpha = 0,
                                      val_frac = 0.20, min_val = 12)
ridge_pred  <- predict(ridge_tuned$fit, s = ridge_tuned$lambda, newx = x_test)
```

### Step 6 --- Fit Lasso (L1 Penalty)

Lasso sets some coefficients exactly to zero, performing automatic variable selection. Useful when only a few lags truly matter (`alpha = 1`).

```{r}
#| label: lasso
lasso_tuned <- tune_glmnet_time_split(x_train, y_train, alpha = 1,
                                      val_frac = 0.20, min_val = 12)
lasso_pred  <- predict(lasso_tuned$fit, s = lasso_tuned$lambda, newx = x_test)

cat("\n--- Lasso Coefficients at chosen lambda ---\n")
print(coef(lasso_tuned$fit, s = lasso_tuned$lambda))
```

### Step 7 --- Compare Out-of-Sample RMSE

```{r}
#| label: rmse-a
calc_rmse <- function(actual, predicted) sqrt(mean((actual - predicted)^2))

results <- data.frame(
  Model = c("OLS — AR(12)", "Ridge — AR(12) (L2)", "Lasso — AR(12) (L1)"),
  RMSE  = c(calc_rmse(y_test, ols_pred),
            calc_rmse(y_test, ridge_pred),
            calc_rmse(y_test, lasso_pred))
)

cat("\n--- Out-of-Sample Performance (RMSE) ---\n")
print(results)
```

### Step 8 --- Plot Out-of-Sample Forecasts

```{r}
#| label: plot-a
#| fig-width: 10
#| fig-height: 5
plot_data <- data.frame(
  Date   = test_dates,
  Actual = y_test,
  OLS    = as.numeric(ols_pred),
  Ridge  = as.numeric(ridge_pred),
  Lasso  = as.numeric(lasso_pred)
)

plot_long <- plot_data %>%
  pivot_longer(cols = c("OLS", "Ridge", "Lasso"),
               names_to = "Model", values_to = "Inflation")

ggplot(plot_long, aes(x = Date)) +
  geom_line(data = plot_data, aes(y = Actual, color = "Actual Data"), linewidth = 1.2) +
  geom_line(aes(y = Inflation, color = Model), linetype = "dashed", linewidth = 0.8) +
  labs(title    = "Out-of-Sample Inflation Forecast: OLS vs. Ridge vs. Lasso",
       subtitle = "US YoY CPI — AR(12) structure with time-ordered tuning",
       x = "Date", y = "Inflation Rate (%)") +
  scale_color_manual(values = c("Actual Data" = "black",
                                "OLS"   = "blue",
                                "Ridge" = "green",
                                "Lasso" = "red")) +
  theme_minimal() +
  theme(legend.title = element_blank(), legend.position = "bottom")
```

------------------------------------------------------------------------

## Part B --- Rolling Windows & h-Step Ahead Forecasting

### Objective

Extend the analysis by evaluating forecasting performance across **multiple horizons** ($h = 1, 3, 6, 12$ months ahead) and **rolling window sizes** (60 and 120 months). This tests whether regularization remains beneficial as the forecasting difficulty increases.

### Key Concepts

**Direct h-step forecasting** fits a separate model for each horizon $h$, predicting $y_{t+h}$ directly from lags at time $t$. This avoids compounding errors from iterated one-step forecasts.

**Rolling windows** retrain the model at each evaluation step using only the most recent $w$ observations, allowing the model to adapt to structural changes over time.

### Setup

```{r}
#| label: setup-b
library(quantmod)
library(glmnet)
library(ggplot2)
library(tidyr)
library(dplyr)

cat("Retrieving US CPI data from FRED...\n")
#getSymbols("CPILFESL", src = "FRED", auto.assign = TRUE)
CPILFESL  <- readRDS("../data/CPILFESL.rds")
inflation <- ROC(CPILFESL, n = 12, type = "discrete") * 100
inflation <- na.omit(inflation)
y_vec <- as.numeric(inflation)
N     <- length(y_vec)
```

### Experiment Parameters

```{r}
#| label: params-b
p        <- 12                   # AR lags
horizons <- c(1, 3, 6, 12)      # Forecast horizons (months ahead)
windows  <- c(60, 120)          # Rolling window sizes: 5 and 10 years
OOS_steps <- 120                # Evaluate on last 10 years of data
```

### Rolling Forecast Loop

For each combination of horizon $h$ and window size $w$, the loop:

1.  Aligns features and targets for direct $h$-step forecasting\
2.  Fits OLS, Ridge, and Lasso on the rolling training window\
3.  Produces a one-step prediction for the current out-of-sample observation

```{r}
#| label: rolling-loop
results_df <- data.frame()

cat("\nStarting Rolling Forecast Experiment...\n")

for (h in horizons) {
  for (w in windows) {
    cat(sprintf(" -> Horizon: %2d month(s) | Window: %3d months\n", h, w))

    preds_ols   <- numeric(OOS_steps)
    preds_ridge <- numeric(OOS_steps)
    preds_lasso <- numeric(OOS_steps)
    actuals     <- numeric(OOS_steps)

    for (step in 1:OOS_steps) {

      i        <- N - OOS_steps + step   # Index being predicted
      origin_t <- i - h                  # Last observation available at forecast time

      # Direct h-step feature alignment
      train_target_idx  <- (origin_t - w + 1):origin_t
      train_feature_idx <- train_target_idx - h

      Y_train <- y_vec[train_target_idx]

      X_train <- matrix(NA, nrow = w, ncol = p)
      for (lag in 1:p) X_train[, lag] <- y_vec[train_feature_idx - lag + 1]

      X_test <- matrix(NA, nrow = 1, ncol = p)
      for (lag in 1:p) X_test[1, lag] <- y_vec[origin_t - lag + 1]

      colnames(X_train) <- colnames(X_test) <- paste0("Lag_", 1:p)

      # OLS
      train_df         <- data.frame(Y = Y_train, X_train)
      preds_ols[step]  <- predict(lm(Y ~ ., data = train_df), data.frame(X_test))

      # Ridge
      suppressWarnings({
        mdl_ridge <- cv.glmnet(X_train, Y_train, alpha = 0)
      })
      preds_ridge[step] <- predict(mdl_ridge, s = "lambda.min", newx = X_test)

      # Lasso
      suppressWarnings({
        mdl_lasso <- cv.glmnet(X_train, Y_train, alpha = 1)
      })
      preds_lasso[step] <- predict(mdl_lasso, s = "lambda.min", newx = X_test)

      actuals[step] <- y_vec[i]
    }

    tmp <- data.frame(
      Horizon_Num = h,
      Window_Num  = w,
      Horizon     = factor(paste0(h, "-Month Ahead"), levels = paste0(horizons, "-Month Ahead")),
      Window      = factor(paste0(w, "M Window"),     levels = paste0(windows,  "M Window")),
      OLS         = sqrt(mean((actuals - preds_ols)^2)),
      Ridge       = sqrt(mean((actuals - preds_ridge)^2)),
      Lasso       = sqrt(mean((actuals - preds_lasso)^2))
    )
    results_df <- rbind(results_df, tmp)
  }
}

cat("\nExperiment Complete!\n")
```

### Results Table

```{r}
#| label: table-b
table_output <- results_df %>%
  select(Window, Horizon, OLS, Ridge, Lasso) %>%
  arrange(Window, Horizon)

print(table_output, row.names = FALSE, digits = 4)
```

### Results Chart

The bar chart below compares RMSE across all horizon and window combinations. Look for patterns: does regularization help more at longer horizons? Does a larger window hurt or help?

```{r}
#| label: plot-b
#| fig-width: 10
#| fig-height: 6
results_long <- results_df %>%
  select(Window, Horizon, OLS, Ridge, Lasso) %>%
  pivot_longer(cols = c(OLS, Ridge, Lasso),
               names_to = "Model", values_to = "RMSE") %>%
  mutate(Model = factor(Model, levels = c("OLS", "Ridge", "Lasso")))

ggplot(results_long, aes(x = Model, y = RMSE, fill = Model)) +
  geom_bar(stat = "identity", position = position_dodge(),
           color = "black", alpha = 0.8) +
  facet_grid(Window ~ Horizon) +
  labs(title    = "Out-of-Sample RMSE: OLS vs. Ridge vs. Lasso",
       subtitle = "Across forecast horizons and rolling window sizes",
       x = "Model", y = "Root Mean Squared Error (RMSE)") +
  scale_fill_manual(values = c("OLS" = "#1f77b4",
                               "Ridge" = "#2ca02c",
                               "Lasso" = "#d62728")) +
  theme_bw() +
  theme(
    plot.title       = element_text(face = "bold", size = 16),
    strip.background = element_rect(fill = "#e9ecef", color = "black"),
    strip.text       = element_text(face = "bold", size = 11),
    axis.text.x      = element_text(angle = 45, hjust = 1, face = "bold"),
    legend.position  = "bottom"
  )
```

------------------------------------------------------------------------

## Discussion Questions

1.  In Part A, which model achieves the lowest out-of-sample RMSE? Does the result surprise you given the number of predictors?
2.  Which Lasso coefficients are set exactly to zero? What does this tell you about the most informative lags?
3.  In Part B, how does RMSE evolve as the horizon $h$ increases? Is the ranking between models stable?
4.  Compare the 60-month vs. 120-month rolling windows. When would a shorter window be preferable?
5.  Why is time-ordered validation (Part A) more appropriate than random k-fold for this type of data?
